---
title: "Posterior Predict"
author: "Alex Dombowsky"
date: '2022-04-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(abind)
library(grid)
library(gridExtra)
```


# Load Data
```{r}
seymour <- read.csv("data/seymour_cd4.csv")
```

## Delete non-continuous variables
```{r}
seymour <- seymour %>% select(-gcs, -Age, -Sex, -Neut)
narows <- apply(seymour, 1, function(x) sum(is.na(x)))
bad <- which(narows>2)
seymour <- seymour[-bad, ]
```

## Get priors
```{r}
priors <- readRDS("priors/new_priors.Rdata")
priors_0 <- readRDS("priors/priors.Rdata")
# add CD4%
priors<- abind(priors, matrix(rep(c(32, 10, 2, 50), 4), byrow = T, nrow = 4, ncol = 4), along = 3)
priors_0 <- abind(priors_0, matrix(rep(c(32, 10, 2, 50), 3), byrow = T, nrow = 3, ncol = 4), along = 3)
```


In which I draw a bunch of predictive samples. Fortunately, we can use the missing data function.
```{r}
source("rfuncts/LCM_Gibbs.R")
fit0 <- readRDS("fits/fit0.Rdata")
```

Can sample using a simple for-loop.
```{r}
R <- (R - burnin) * n_chains
B <- 0
stop_ind <- 500
stops <- (1:(R/stop_ind)) * stop_ind
n <- nrow(seymour)
p <- ncol(seymour)
pps <- array(0, dim = c(R-B, n, p))
theta <- fit0$theta
s <- fit0$s
for (r in 1:R){
  for (i in 1:n){
    for (j in 1:p) {
      pps[r, i, j] <- sample.y_ij(theta_j = theta[[j]][, ,r], type = "c", s_i = s[r, i])
    }
  }
      # print stops
  if (r %in% stops){
    print(r)
  }
}
saveRDS(pps, "fits/posterior_predictive.Rdata")
pps <- readRDS("fits/posterior_predictive.Rdata")

```

# Idea 1: Distributions in Each Cluster

## test out on just hemoglobin
```{r}
y <- seymour
K <- 3
y$cluster <- factor(s0.minbinder$cl, levels = 1:K)
plots <- list()
for (h in 1:K) {
  test = y %>% select(hgb, cluster) %>% filter(cluster == h) %>% select(-cluster)
  plots[[h]] <- ggplot(test, aes(x = hgb)) + geom_histogram(color = "blue", fill = "skyblue1") +
  labs(title = toString(h)) + xlim(4,22)
}
grid.arrange(plots[[1]], plots[[2]],
             plots[[3]],
             nrow = 2, ncol = 2)

```

## write function
```{r}
cluster_summarize <- function(data, clustering, K, names) {
  p <- ncol(data)
  colnames(data) <- names
  data$cluster <- factor(clustering, levels = 1:K)
  plots <- list()
  for (j in 1:p){
    test = data %>% select(names[j], cluster)
    colnames(test) = c("var", "cluster")
    plot <- test %>% ggplot(aes(x = var, color = cluster, fill = cluster)) + geom_density(alpha = 0.5) + labs(title = names[j]) + xlab("Variable") +
      theme(text = element_text(size = 13)) + theme_bw()
    print(plot)
    #for (h in 1:K) {
      #test = y %>% select(names[j], cluster) %>% filter(cluster == h) %>% select(-cluster)
      #colnames(test) = "var"
      #plots[[h]] <- ggplot(test, aes(x = var)) + geom_density(color = "blue", fill = "skyblue1") +
       #xlim(min(data[,j], na.rm = T), max(data[,j], na.rm = T)) + 
        #labs(title = paste0("h=",toString(h)))
    #}
    #grid.arrange(plots[[1]], plots[[2]],
             #plots[[3]], plots[[4]],
             #nrow = 2, ncol = 2,
             #top = textGrob(names[j],gp=gpar(fontsize=20,font=3)))
  }
}

```

## implement function
```{r, warning = F}
names <- c("Hemoglobin", "Platelet Count", "Pulse Rate", "Blood Pressure",
           "Lactate", "Respiratory Rate", "Oxygen Saturation", "Temperature",
           "White Blood Cell Count", "CD4 Percent")
cluster_summarize(data = seymour, clustering = s0.minbinder$cl, K = 3, names = names)
```
# now, do function with posterior predictive samples
```{r}
# can look at how vars are separated at each Gibbs iteration
r = 100
cluster_summarize(data = as.data.frame(pps[r,,]), clustering = s[r,], K = 4,
                  names = names)


```
# Prior Predictive vs. Posterior Predictive
```{r}
# sample from prior predictive
source("rfuncts/prior_predictive.R")
pri.pred <- prior_predict(n = nrow(seymour), p = ncol(seymour), alpha = rep(1/3,3),
                          priors = priors_0, names = colnames(seymour))
y_prior <- pri.pred$y
s_prior <- pri.pred$s

# plot using above function
cluster_summarize(data = y_prior, clustering = s_prior, K = 4,
                  names = names)
```



